\section{Introduction}

The Ising model or Lenz-Ising \cite{Ising1925}, named after german physicists
Ernst Ising and Wilhelm Lenz, is a mathematical model for ferromagnetism in
statistical mechanics. It consists of discrete variables which represent the
magnetic dipole moments of atomic "spins" and  can take on values of either
+1 or -1. These spins are arranged on a lattice, and the model defines the
interactions between these spins.

This lattice structure repeats periodically in all directions, allowing each
spin to interact with its neighboring spins. The key aspect of the model is that
neighboring spins with the same orientation (either both +1 or both -1) have
lower energy than those with opposing orientations. This setup allows the model
to capture the fundamental behavior of ferromagnetic materials.

One of the intriguing aspects of the Ising model is the way it balances energy
minimization and thermal disturbance \cite{Chowdhury}. While the system
naturally tends towards a state of lowest energy, thermal fluctuations can
disrupt this tendency, leading to two structural phases namely the paramagnetic
phase and the ferromagnetic phase. The paramagnetic phase is characterized by a
random orientation of spins and the ferromagnetic phase is characterized by a
uniform orientation of spins. The transition between these two phases is known
as the phase transition. The Ising model is one of the simplest models to
exhibit a phase transition however, it is still a non-trivial problem to solve
analytically. It is important to note that 1D Ising model does not exhibit a
phase transition. The phase transition occurs only in 2D \cite{Onsager1944} and
higher dimensions. This dynamic is crucial for understanding phase transitions
in physical systems.

The concept of the Ising Model dates back to 1920 when Wilhelm Lenz proposed it,
later tackled by his student Ernst Ising. In his 1924 thesis \cite{Ising1925},
Ising solved the one-dimensional version of the model, demonstrating the absence
of a phase transition in this case. The more complex two-dimensional
square-lattice version remained unsolved until 1944, when Lars Onsager provided
a groundbreaking analytic solution \cite{Onsager1944}, marking a significant
advancement in statistical mechanics.
